# The Family of Autoregressive Moving Average Models

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(simts)
library(ggplot2)
library(robcor)
```

In this chapter we introduce a class of time series models that is considerably flexible and among the most commonly used to describe stationary time series. This class is represented by the Seasonal AutoRegressive Integrated Moving Average (SARIMA) models which, among others, combine and include the autoregressive and moving average models seen in the previous chapter. To introduce this class of models, we start by describing a sub-class called AutoRegressive Moving Average (ARMA) models which represent the backbone on which the SARIMA class is built. The importance of ARMA models resides in their flexibility as well as their capacity of describing (or closely approximating) almost all the features of a stationary time series. The autoregressive parts of these models describe how consecutive observations in time influence each other while the moving average parts capture some possible unobserved shocks thereby allowing to model different phenomena which can be observed in various fields going from biology to finance.

With this premise, the first part of this chapter introduces and explains the class of ARMA models in the following manner. First of all we will discuss the class of linear processes, which ARMA models belong to, and we will then proceed to a detailed description of autoregressive models in which we review their definition, explain their properties, introduce the main estimation methods for their parameters and highlight the diagnostic tools which can help understand if the estimated models appear to be appropriate or sufficient to well describe the observed time series. Once this is done, we will then use most of the results given for the autoregressive models to further describe and discuss moving average models, for which we underline the property of invertibility, and finally the ARMA models. Indeed, the properties and estimation methods for the latter class are directly inherited from the discussions on the autoregressive and moving average models.

The second part of this chapter introduces the general class of SARIMA models, passing through the class of ARIMA models. These models allow to apply the ARMA modeling framework also to time series that have particular non-stationary components to them such as, for example, linear and/or seasonal trends. Extending ARMA modeling to these cases allows SARIMA models to be an extremely flexible class of models that can be used to describe a wide range of phenomena.


## Linear Processes

In order to discuss the classes of models mentioned above, we first present the class of linear processes which underlie many of the most common time series models.

```{definition, label="lp", name="Linear Process"}
A time series, $(X_t)$, is defined to be a linear process if it can be expressed
as a linear combination of white noise as follows:

\[{X_t} = \mu + \sum\limits_{j =  - \infty }^\infty  {{\psi _j}{W_{t - j}}} \]

where $W_t \sim WN(0, \sigma^2)$ and $\sum\limits_{j =  - \infty }^\infty  {\left| {{\psi _j}} \right|}  < \infty$.
```

Note, the latter assumption is required to ensure that the series has a limit. Furthermore, the set of coefficients \[{\{ {\psi _j}\} _{j =  - \infty , \cdots ,\infty }}\] can be viewed as a linear filter. These coefficients do not have to be all equal nor symmetric as later examples will show. Generally, the properties of a linear process related to mean and variance are given by:

\[\begin{aligned}
\mu_{X} &= \mu \\
\gamma_{X}(h) &= \sigma _W^2\sum\limits_{j =  - \infty }^\infty  {{\psi _j}{\psi _{h + j}}}
\end{aligned}\]

The latter is derived from 

\[\begin{aligned}
  \gamma \left( h \right) &= Cov\left( {{x_t},{x_{t + h}}} \right) \\
   &= Cov\left( {\mu  + \sum\limits_{j =  - \infty }^\infty  {{\psi _j}{w_{t - j}}} ,\mu  + \sum\limits_{j =  - \infty }^\infty  {{\psi _j}{w_{t + h - j}}} } \right) \\
   &= Cov\left( {\sum\limits_{j =  - \infty }^\infty  {{\psi _j}{w_{t - j}}} ,\sum\limits_{j =  - \infty }^\infty  {{\psi _j}{w_{t + h - j}}} } \right) \\
   &= \sum\limits_{j =  - \infty }^\infty  {{\psi _j}{\psi _{j + h}}Cov\left( {{w_{t - j}},{w_{t - j}}} \right)}  \\
   &= \sigma _w^2\sum\limits_{j =  - \infty }^\infty  {{\psi _j}{\psi _{j + h}}}  \\ 
\end{aligned} \]

Within the above derivation, the key is to realize that 
$Cov\left( {{w_{t - j}},{w_{t + h - j}}} \right) = 0$ if $t - j \ne t + h - j$.

Lastly, another convenient way to formalize the definition of a linear process is through the use of the **backshift operator** (or lag operator) which is itself defined as follows:

\[B\,X_t = X_{t-1}.\]

The properties of the backshift operator allow us to create composite functions of the type 

$$B^2 \, X_t = B (B \, X_t) = B \, X_{t-1} = X_{t-2}$$
which allows to generalize as follows

$$B^k \, X_t = X_{t-k}.$$
Moreover, we can apply the inverse operator to it (i.e. $B^{-1} \, B = 1$) thereby allowing us to have, for example:

$$X_t = B^{-1} \, B X_t = B^{-1} X_{t-1}$$

```{example, label="backdiff", name="d-order Differences"}
We can re-express $X_t - X_{t-1}$ as
$$\delta X_t = (1 - B) X_t$$
or a second order difference as
$$\delta^2 X_t = (1 - B)^2 X_t$$
thereby generalizing to a d-order difference as follows:
$$\delta^d X_t = (1 - B)^d X_t.$$
```

Having defined the backshift operator, we can now provide an alternative definition of a linear process as follows:

\[{X_t} = \psi \left( B \right){W_t}\]

where $\psi ( B )$ is a polynomial function in $B$ whose coefficients are given by the linear filters $\{\psi_j\}$ (we'll describe these polynomials further on).

```{example, label="lpwn", name="Linear Process of White Noise"}

The white noise process $\left\{X_t\right\}$, defined in \@ref(wn),
can be expressed as a linear process as follows:

\[\psi _j = \begin{cases}
      1 , &\mbox{ if } j = 0\\
      0 , &\mbox{ if } |j| \ge 1
\end{cases}.\]

and $\mu = 0$.

Therefore, $X_t = W_t$, where $W_t \sim WN(0, \sigma^2_W)$
```

```{example, label="lpma1", name="Linear Process of Moving Average Order 1"}

Similarly, consider $\left\{X_t\right\}$ to be a MA(1) process, 
given by \@ref(ma1). The process can be expressed linearly through the following filters: 

\[\psi _j = \begin{cases}
      1, &\mbox{ if } j = 0\\
      \theta , &\mbox{ if } j = 1 \\
      0, &\mbox{ if } j \ge 2
\end{cases}.\]

and $\mu = 0$.

Thus, we have: $X_t = W_t + \theta W_{t-1}$
```

```{example, label="lpsma", name="Linear Process and Symmetric Moving Average"}

Consider a symmetric moving average given by:

\[{X_t} = \frac{1}{{2q + 1}}\sum\limits_{j =  - q}^q {{W_{t + j}}} \]

Thus, $\left\{X_t\right\}$ is defined for $q + 1 \le t \le n-q$. The above process
would be a linear process since:

\[\psi _j = \begin{cases}
      \frac{1}{{2q + 1}} , &\mbox{ if } -q \le j \le q\\
      0 , &\mbox{ if } |j| > q
\end{cases}.\]

and $\mu = 0$.

In practice, if $q = 1$, we would have:

\[{X_t} = \frac{1}{3}\left( {{W_{t - 1}} + {W_t} + {W_{t + 1}}} \right)\]
```


```{example, label="lpar1", name="Autoregressive Process of Order 1"}
If $\left\{X_t\right\}$ follows an AR(1) model defined in \@ref(ar1), the linear filters are a function of the time lag:

\[\psi _j = \begin{cases}
      \phi^j , &\mbox{ if } j \ge 0\\
      0 , &\mbox{ if } j < 0
\end{cases}.\]

and $\mu = 0$. We would require the condition that $\left| \phi \right| < 1$ in order to respect the condition on the filters (i.e. $\sum\limits_{j =  - \infty }^\infty  {\left| {{\psi _j}} \right|}  < \infty$).
```


## Autoregressive Models (AR(p))

The class of autoregressive models is based on the idea that previous values in the time series are needed to explain current values in the series. For this class of models, we assume that the $p$ previous observations are needed for this purpose and we therefore denote this class as AR($p$). In the previous chapter, the model we introduced was an AR(1) in which only the immediately previous observation is needed to explain the following one and therefore represents a particular model which is part of the more general class of AR(p) models.

```{definition, label="ar_p","Autoregressive Models of Order P"}
The AR(p) models can be formally represented as follows
$${X_t} = {\phi_1}{Y_{t - 1}} + ... + {\phi_p}{X_{t - p}} + {W_t},$$
where $\phi_i \neq 0$ (for $i = 1, ..., p$) and $W_t$ is a (Gaussian) white noise process with variance $\sigma^2$. 
```

As earlier in this book, we will assume that the expectation of the process $({X_t})$, as well as that of the following ones in this chapter, is zero. The reason for this simplification is that if $\mathbb{E} [ X_t ] = \mu$, we can define an AR process *around* $\mu$ as follows:

$$X_t - \mu = \sum_{i = 1}^p \phi_i \left(X_{t-i} - \mu \right) + W_t,$$

which is equivalent to 

$$X_t  = \mu^{\star} +  \sum_{i = 1}^p \phi_i X_{t-i}  + W_t,$$

where $\mu^{\star} = \mu (1 - \sum_{i = 1}^p \phi_i)$. Therefore, to simplify the notation we will generally consider only zero mean processes, since adding means (as well as other deterministic trends) is easy.

A useful way of representing AR(p) processes is through the backshift operator introduced in the previous section and is as follows

\[\begin{aligned}
  {X_t} &= {\phi_1}{X_{t - 1}} + ... + {\phi_p}{y_{t - p}} + {w_t} \\
   &= {\phi_1}B{X_t} + ... + {\phi_p}B^p{X_t} + {W_t} \\
   &= ({\phi_1}B + ... + {\phi_p}B^p){X_t} + {W_t} \\ 
\end{aligned},\]

which finally yields

$$(1 - {\phi _1}B - ... - {\phi_p}B^p){X_t} = {W_t},$$

which, in abbreviated form, can be expressed as

$$\phi(B){X_t} = W_t.$$

We will see that $\phi(B)$ is important to establish the stationarity of these processes and is called the *autoregressive* operator. Moreover, this quantity is closely related to another important property of AR(p) processes called *causality*. Before formally defining this new property we consider the following example which provides an intuitive illustration of its importance.

**Example:** Consider a classical AR(1) model with $|\phi| > 1$. Such a model could be expressed as

$$X_t = \phi^{-1} X_{t+1} - \phi^{-1} W_t = \phi^{-k} X_{t+k} - \sum_{i = 1}^{k-1} \phi^{-i} W_{t+i}.$$

Since $|\phi| > 1$, we obtain

$$X_t = - \sum_{j = 1}^{\infty} \phi^{-j} W_{t+j},$$

which is a linear process and therefore is stationary. Unfortunately, such a model is useless because we need the future to predict the future. These processes are called non-causal.

### Properties of AR(p) models

In this section we will describe the main property of the AR(p) model which has already been mentioned in the previous paragraphs and therefore let us now introduce the property of causality in a more formal manner.

**Definition:** An AR(p) model is *causal* if the time series $\{ X_t \}_{-\infty}^{\infty}$ can be written as a one-sided linear process:
\begin{equation}
    X_t = \sum_{j = 0}^{\infty} \psi_j W_{t-j} = \frac{1}{\phi(B)} W_t = \psi(B) W_t,
(\#eq:causal)
\end{equation}
where $\phi(B) = \sum_{j = 0}^{\infty} \phi_j B^j$, and $\sum_{j=0}^{\infty}|\phi_j| < \infty$ and setting $\phi_0 = 1$.

As discussed earlier this condition implies that only the past values of the time series can explain the future values of it and not viceversa. Moreover, given the expression of the linear filters given by
$$\frac{1}{\phi(B)}$$
it is obvious that a solution exists only when $\phi(B) = \sum_{j = 0}^{\infty} \phi_j B^j \neq 0$ (thereby implying causality). A condition for this to be respected is for the roots of $\phi(B) = 0$ to lie outside the unit circle.

<!-- However, it might be difficult and not obvious to show the causality of an AR(p) process by using the above definitions directly, thus the following properties are useful in practice.  -->

<!-- **Causality** -->
<!-- If an AR(p) model is causal, then the coefficients of the one-sided linear process given in \@ref(eq:causal) can be obtained by solving -->
<!-- \begin{equation*} -->
<!--     \psi(z) = \frac{1}{\sum_{j=0}^{\infty} \phi_j z^j} = \frac{1}{\phi(z)}, \mbox{ } |z| \leq 1. -->
<!-- \end{equation*} -->

<!-- It can be seen how there is no solution to the above equation if $\phi(z) = 0$ and therefore an AR(p) is causal if and only if $\phi(z) \neq 0$. A condition for this to be respected is for the roots of $\phi(z) = 0$ to lie outside the unit circle. -->


## Estimation of AR(p) models

Given the above defined properties of AR(p) models, we will now discuss how these models can be estimated, more specifically how the $p+1$ parameters can be obtained from an observed time series. Indeed, a reliable estimation of these models is necessary in order to intepret and describe different natural phenomena and/or forecast possible future values of the time series.

A first approach builds upon the earlier definition of AR(p) models being a linear process. Recall that
\begin{equation}
    X_t = \sum_{j = 1}^{p} \phi_j X_{t-j}
\end{equation}
which delivers the following autocovariance function
\begin{equation}
    \gamma(h) = \text{cov}(X_{t+h}, X_t) = \text{cov}\left(\sum_{j = 1}^{p} \phi_j X_{t-j}, X_t\right) = \sum_{j = 1}^{p} \phi_j \gamma(h-j), \mbox{ } h \geq 1.
\end{equation}
Rearranging the above expressions we obtain the following general equations
\begin{equation}
    \gamma(h) - \sum_{j = 1}^{p} \phi_j \gamma(h-j) = 0, \mbox{ } h \geq 1
\end{equation}
and, recalling that $\gamma(h) = \gamma(-h)$,
\begin{equation}
    \gamma(0) - \sum_{j = 1}^{p} \phi_j \gamma(j) = \sigma_w^2.
\end{equation}
We can now define the Yule-Walker equations.

**Definition:** The Yule-Walker equations are given by
\begin{equation}
    \gamma(h) = \phi_1 \gamma(h-1) + ... + \phi_p \gamma(h-p), \mbox{ } h = 1,...,p
\end{equation}
and
\begin{equation}
    \sigma_w^2 = \gamma(0) - \phi_1 \gamma(1) - ... - \phi_p \gamma(p).
\end{equation}
which in matrix notation can be defined as follows
\begin{equation}
    \Gamma_p \mathbf{\phi} = \mathbf{\gamma}_p \,\, \text{and} \,\, \sigma_w^2 = \gamma(0) - \mathbf{\phi}'\mathbf{\gamma}_p
\end{equation}
where $\Gamma_p$ is the $p\times p$ matrix containing the autocovariances $\gamma(k-j)$, where $j,k = 1, ...,p$, while $\mathbf{\phi} = (\phi_1,...,\phi_p)'$ and $\mathbf{\gamma}_p = (\gamma(1),...,\gamma(p))'$ are $p\times 1$ vectors.

Considering the Yule-Walker equations, it is possible to use a method of moments approach and simply replace the theoretical quantities given in the previous definition with their empirical (estimated) counterparts that we saw in the previous chapter. This gives us the following Yule-Walker estimators
\begin{equation}
    \hat{\mathbf{\phi}} = \hat{\Gamma}_p^{-1}\hat{\mathbf{\gamma}}_p \,\, \text{and} \,\, \hat{\sigma}_w^2 = \hat{\gamma}(0) - \hat{\mathbf{\gamma}}_p'\hat{\Gamma}_p^{-1}\hat{\mathbf{\gamma}}_p .
\end{equation}

These estimators have the following asymptotic properties.

**Consistency and Asymptotic Normality of Yule-Walker estimators:**
The Yule-Walker estimators for a causal AR(p) model have the following asymptotic properties:

\begin{equation*}
\sqrt{T}(\hat{\mathbf{\phi}}- \mathbf{\phi}) \xrightarrow{\mathcal{D}} \mathcal{N}(\mathbf{0},\sigma_w^2\Gamma_p^{-1}) \,\, \text{and} \,\, \hat{\sigma}_w^2 \xrightarrow{\mathcal{P}} \sigma_w^2 .
\end{equation*}

Therefore the Yule-Walker estimators have an asymptotically normal distribution and the estimator of the innovation variance is consistent. Moreover, these estimators are also optimal for AR(p) models, meaning that they are also efficient. However, there is also another method which allows to achieve this efficiency (also for general ARMA models that will be tackled further on) and this is the Maximum Likelihood (ML) method. Considering an AR(1) model as an example, and assuming without loss of generality that its expectation is zero, we have the following representation of the AR(1) model
\begin{equation*}
X_t = \phi X_{t-1} + w_t
\end{equation*}
where $|\phi|<1$ and $w_t \overset{iid}{\sim} \mathcal{N}(0,\sigma_w^2)$. Supposing we have observations $(x_t)_{t=1,...,T}$ issued from this model, then the likelihood function for this setting is given by
\begin{equation*}
L(\phi,\sigma_w^2) = f(\phi,\sigma_w^2|x_1,...,x_T)
\end{equation*}
which, for an AR(1) model, can be rewritten as follows
\begin{equation*}
L(\phi,\sigma_w^2) = f(x_1)f(x_2|x_1)\cdot \cdot \cdot f(x_T|x_{T-1}).
\end{equation*}
If we define $\Omega_t^p$ as the information contained in the previous $p$ observations (before time $t$), the above expression can be generalized for an AR(p) model as follows
\begin{equation*}
L(\phi,\sigma_w^2) = f(x_1,...,x_p)f(x_{p+1}|\Omega_{p+1}^p)\cdot \cdot \cdot f(x_T|\Omega_{T-1}^p)
\end{equation*}
where $f(x_1,...,x_p)$ is the joint probability distribution of the first $p$ observations. Going back to the AR(1) setting, based on our assumption on $(w_t)$ we know that $x_t|x_{t-1} \sim \mathcal{N}(\phi x_{t-1},\sigma_w^2)$ and therefore we have that
\begin{equation*}
f(x_t|x_{t-1}) = f_w(x_t - \phi x_{t-1})
\end{equation*}
where $f_w(\cdot)$ is the distribution of $w_t$. This rearranges the likelihood function as follows
\begin{equation*}
L(\phi,\sigma_w^2) = f(x_1)\prod_{t=2}^T f_w(x_t - \phi x_{t-1})
\end{equation*}
where $f(x_1)$ can be found through the causal representation
\begin{equation*}
x_1 = \sum_{j=0}^{\infty} \phi^j w_{1-j} 
\end{equation*}
which implies that $x_1$ follows a normal distribution with zero expectation and a variance given by $\frac{\sigma_w^2}{(1-\phi^2)}$. Based on this, the likelihood function of an AR(1) finally becomes
\begin{equation*}
L(\phi,\sigma_w^2) = (2\pi \sigma_w^2)^{-\frac{T}{2}} (1 - \phi)^{\frac{1}{2}} \exp \left(-\frac{S(\phi)}{2 \sigma_w^2}\right)
\end{equation*}
with $S(\phi) = (1-\phi)^2 x_1^2 + \sum_{t=2}^T (x_t -\phi x_{t-1})^2$. Once the derivative of the logarithm of the likelihood is taken, the minimization of the negative of this function is usually done numerically. However, if we condition on the initial values, the AR(p) models are linear and, for example, we can then define the conditional likelihood of an AR(1) as
\begin{equation*}
L(\phi,\sigma_w^2|x_1) = (2\pi \sigma_w^2)^{-\frac{T-1}{2}} \exp \left(-\frac{S_c(\phi)}{2 \sigma_w^2}\right)
\end{equation*}
where
\begin{equation*}
S_c(\phi) = \sum_{t=2}^T (x_t -\phi x_{t-1})^2 .
\end{equation*}
The latter is called the conditional sum of squares and $\phi$ can be estimated as a straightforward linear regression problem. Once an estimate $\hat{\phi}$ is obtained, this can be used to obtain the conditional maximum likelihood estimate of $\sigma_w^2$
\begin{equation*}
\hat{\sigma}_w^2 = \frac{S_c(\hat{\phi})}{(T-1)} .
\end{equation*}

The estimation methods presented so far are standard for these kind of models. Nevertheless, if the data suffers from some form of contamination, these methods can become highly biased. For this reason, some robust estimators are available to limit this problematic if there are indeed outliers in the observed time series. One of these methods relies on the estimator proposed in Kunsch (1984) who underlines that the MLE score function of an AR(p) is given by
\begin{equation*}
 \kappa(\mathbf{\theta}|x_j,...x_{j+p}) = \frac{\partial}{\partial \mathbf{\theta}} (x_{j+p} - \sum_{k=1}^p \phi_k x_{j+p-k})^2
\end{equation*}
where $\theta$ is the parameter vector containing, in the case of an AR(1)
model, the two parameters $\phi$ and $\sigma_w^2$ 
(i.e. $\theta = [\phi \,\, \sigma_w^2]$). This delivers the estimating equation
\begin{equation*}
\sum_{j=1}^{n-p} \kappa (\hat{\mathbf{\theta}}|x_j,...x_{j+p}) = 0 .
\end{equation*}
The score function $\kappa(\cdot)$ is clearly not bounded, in the sense that if
we arbitrarily move a value of $(x_t)$ to infinity then the score function also 
goes to infinity thereby delivering a biased estimation procedure. To avoid that
outlying observations bias the estimation excessively, a bounded score function
can be used to deliver an M-estimator given by
\begin{equation*}
\sum_{j=1}^{n-p} \psi (\hat{\mathbf{\theta}}|x_j,...x_{j+p}) = 0,
\end{equation*}
where $\psi(\cdot)$ is a function of bounded variation. When conditioning on the
first $p$ observations, this problem can be brought back to a linear regression
problem which can be applied in a robust manner using the robust regression
tools available in `R` such as `rlm(...)` or `lmrob(...)`. However, another 
available tool in `R` which is does not require a strict specification of the distribution function (also for general ARMA models) is the `gmwm(...)` function in which it is possible to specify the option `robust = TRUE`. This function makes use of a quantity called the wavelet variance (denoted as $\nu$) which is estimated robustly and then used to retrieve the parameters $\theta$ of the time series model. The robust estimate is obtained by solving the following minimization problem
\begin{equation*}
\hat{\boldsymbol{\theta}} = \underset{\boldsymbol{\theta} \in \boldsymbol{\Theta}}{\text{argmin}} (\hat{\boldsymbol{\nu}} - \boldsymbol{\nu}(\boldsymbol{\theta}))^T\boldsymbol{\Omega}(\hat{\boldsymbol{\nu}} - \boldsymbol{\nu}({\boldsymbol{\theta}})),
\end{equation*}
where $\hat{\nu}$ is the robustly estimated wavelet variance, $\nu(\theta)$ is 
the theoretical wavelet variance (implied by the model we want to estimate) and $\Omega$ is a positive definite weighting matrix. Below we show some simulation studies where we present the results of the above estimation procedures in absence and in presence of contamination in the data.

In particular, we simulate three different 
processes $X_t, Y_t, Z_t$ from \[X_t = 0.5 X_{t-1} - 0.25 X_{t-2} + W_t\]
with $W_t \overset{iid}{\sim} N(0, 1)$. The first process $(X_t)$ is uncontaminated while the other two processes are contaminated versions of the first that can often be observed in practice. The first type of contamination can be seen in $(Y_t)$ and is delivered by replacing a portion of the original process with a process defined as \[U_t = 0.90 U_{t-1} - 0.40 U_{t-2} + V_t\], where $V_t \overset{iid}{\sim} N(0, 9)$. The second form of contamination can be seen in $(Z_t)$ and consists in the so-called point-wise contamination where randomly selected points from $X_t$ are replaced with $N_t \overset{iid}{\sim} N(0, 9)$.


``` {r simuAR2data, cache = TRUE}
n         = 1000           # Sample Size (T)
eps       = 0.05           # Proportion of contamination
cont      = round(eps*n) # Number of contaminated observations

set.seed(19)               # Set seed for reproducibility

# Generate time series
Xt = gen_gts(n, AR(phi = c(0.5,0.25), sigma2 = 1))
Yt = gen_gts(n, AR(phi = c(0.5,0.25), sigma2 = 1))
Zt = gen_gts(n, AR(phi = c(0.5,0.25), sigma2 = 1))

# Contaminate a portion of Yt with a process
index = sample(1:(n-cont), 1) 
Yt[101:135,] = gen_gts(35, AR(phi = c(0.9,-0.4), sigma2 = 9))

# Contaminate at random Zt with noise
Zt = Yt
#Zt[sample(n, dispersal, replace = FALSE),] = #gen_gts(dispersal, WN(sigma2 = 9))
```

```{r visualizesimuAR2data, cache = TRUE, echo = FALSE, fig.height = 9, fig.width = 6, fig.cap = "Contaminated AR(2) Processes", fig.align='center'}
# Graph points
par(mfrow = c(3,1))
plot(Xt, main = expression(paste("Simulated process ",X[t])))
plot(Yt, main = expression(paste("Simulated process ",Y[t])))
plot(Zt, main = expression(paste("Simulated process ",Z[t])))
```

As every process will require the same estimation methodology, we will first define a computational engine that will compute estimates for the parameters across all of the different methods under the simulation study. Doing so provides separate, reusable code that lowers the threshold for error when compared to having multiple copies of the same code scattered about. In addition, we have defined a means to visualize the results obtained from the simulation study through a boxplot. The procedure for creating the graphs is detailed after the simulation engine.



```{r simengine, cache = TRUE}
sim_engine = function(Xt){
  # Estimate ARIMA parameters using MLE
  mod = arima(Xt, order = c(2, 0, 0), include.mean  = FALSE)
  
  # Extract MLE Parameters (including sigma)
  res.MLE = c(mod$coef, mod$sigma)
  
  # Calculate ACF
  autocorr = as.numeric(acf(Xt, lag.max = 2, plot = FALSE)$acf)
  X = matrix(1,2,2)
  X[1,2] = X[2,1] = autocorr[2]
  y = autocorr[2:3]
  
  # Compute Least Squares on ACF
  svmat = solve(X)
  phi.LS = svmat%*%y
  sig2.LS = var(Xt)*(1 - t(y)%*%svmat%*%y)
  res.LS = c(phi.LS, sig2.LS)
  
  # Calculate RACF
  rob.ccf = as.numeric(robcor::robacf(Xt, plot=FALSE, type = "covariance")$acf)
  X[1,2] = X[2,1] = rob.ccf[2]/rob.ccf[1]
  y = rob.ccf[2:3]/rob.ccf[1]
  
  # Compute Least Squares on RACF
  svmat = solve(X)
  phi.RR = svmat%*%y
  sig2.RR = rob.ccf[1]*(1 - t(y)%*%svmat%*%y)
  res.RR = c(phi.RR, sig2.RR)
  
  # Compute the GMWM Estimator
  res.GMWM = gmwm2::gmwm(ARMA(2,0), Xt)$estimate
  res.RGMWM = gmwm2::gmwm(ARMA(2,0), Xt, robust = TRUE)$estimate

  # Return results
  list(res.MLE = res.MLE, res.LS = res.LS, res.RR = res.RR,
       res.GMWM = res.GMWM, res.RGMWM = res.RGMWM)
} 
```

```{r simgraphcreator, cache = TRUE}
sim_study_graph = function(res.MLE, res.LS, res.RR, res.GMWM, res.RGMWM,
                           Truth = c(0.5, 0.25, 1)){
  # Converts Simulation Matrices to Long Form for use in ggplot2
  d = bmisc::study_df(res.MLE, res.LS, res.RR, res.GMWM, res.RGMWM,
                         data_names = c("MLE","LS","RLS","GMWM","RGMWM"))
  
  # Reorder factors
  d$Draw = factor(d$Draw, levels = c(1,2,3),
                  labels = expression(phi[1], phi[2], sigma^2))
  d$Type = factor(d$Type, levels = c("MLE","LS","RLS","GMWM","RGMWM"))
  
  # Add in horizontal lines
  oracle = data.frame(Draw = 1:3, Truth = Truth)
  oracle$Draw = factor(oracle$Draw, levels = c(1,2,3), 
                       labels = expression(phi[1], phi[2], sigma^2))
  
  # Plot
  ggplot(d, aes(x = Type, y = Values, fill = Type)) + 
    geom_boxplot() + 
    stat_boxplot(geom ='errorbar', width = 0.5) + 
    geom_hline(data = oracle, aes(yintercept = Truth), color = "red") +
    facet_wrap(~Draw, labeller = label_parsed, scales = "free_y") +
    theme_bw() + 
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 50, hjust = 1))
}
```

With the means to now compute and visualize parameter estimates for each of the given methods, the bootrap simulation study may now commence. Using the
simulation engine, we can simultaneously evaluate the processes within one
loop iteration. Please take into consideration that the following simulation study is computationally intensive and may require the amount of iteration $B$ to be decreased on a personal computer to complete in a reasonable time.


``` {r simuAR2study, eval = FALSE, cache = TRUE, fig.height = 4.5, fig.width = 9}
# Number of bootstrap iterations
B = 250

# Simulation storage
res.xt.MLE = res.xt.LS = res.xt.RR = res.xt.GMWM = res.xt.RGMWM = matrix(NA, B, 3)
res.yt.MLE = res.yt.LS = res.yt.RR = res.yt.GMWM = res.yt.RGMWM = matrix(NA, B, 3)
res.zt.MLE = res.zt.LS = res.zt.RR = res.zt.GMWM = res.zt.RGMWM = matrix(NA, B, 3)

# Begin bootstrap
for (i in seq_len(B)){
  
  # Set seed for reproducibility
  set.seed(i)
  
  # Generate processes
  Xt = gen_gts(n, AR(phi = c(0.5, 0.25), sigma2 = 1))
  Yt = gen_gts(n, AR(phi = c(0.5, 0.25), sigma2 = 1))
  Zt = gen_gts(n, AR(phi = c(0.5, 0.25), sigma2 = 1))
  
  # Generate Ut contamination process that replaces a portion of original signal
  Yt[101:135] = gen_gts(35, AR(phi = c(0.9,-0.4), sigma2 = 9))
  
  # Generate Nt contamination that inject noise at random
  Zt[sample(n, dispersal, replace = FALSE)] = gen_gts(dispersal, WN(sigma2 = 9))

  # Compute estimates and store in the appropriate matrix
  res = sim_engine(Xt)
  res.xt.MLE[i,]  = res$res.MLE; res.xt.LS[i,] = res$res.LS; res.xt.RR[i,] = res$res.RR
  res.xt.GMWM[i,] = res$res.GMWM;res.xt.RGMWM[i,] = res$res.RGMWM
  
  res = sim_engine(Yt)
  res.yt.MLE[i,]  = res$res.MLE; res.yt.LS[i,] = res$res.LS; res.yt.RR[i,] = res$res.RR
  res.yt.GMWM[i,] = res$res.GMWM;res.yt.RGMWM[i,] = res$res.RGMWM

  res = sim_engine(Zt)
  res.zt.MLE[i,]  = res$res.MLE; res.zt.LS[i,] = res$res.LS; res.zt.RR[i,] = res$res.RR
  res.zt.GMWM[i,] = res$res.GMWM;res.zt.RGMWM[i,] = res$res.RGMWM
}
```

