# Robust Regression Methods {#appendixb}

This appendix is largely based on the introduction to linear robust regression
presented in @ronchetti2006historical and @duncan2016ela. In these references
it is stated that the vast majority of the statistical models employed in 
different fields going from finance to biology and engineering, for example,
are parametric models. Based on these models, assumptions are made concerning
the properties of the variables of interest (and the models themselves) and
optimal procedures are derived under these assumptions. Among these procedures,
the least squares and maximum likelihood estimators are well known examples
that, however, are only optimal when the underlying statistical assumptions are
exactly satisfied. If the latter case does not hold, then these procedures can
become considerably biased and/or inefficient when there exist small deviations
from the model. The results obtained by classical procedures can therefore be
misleading when applied to real data (see e.g. @ronchetti2006historical and
@huber2009robust).

In order to address the problems arising from violated parametric assumptions,
robust statistics can be seen as an extension to classical parametric statistics
by directly considering the deviations from the models. Indeed, while parametric
models may be a good approximation of the true underlying situation, robust
statistics does not assume that the model is exactly correct. A robust procedure
as stated in @huber2009robust therefore should have the following features: 

- It should efficiently estimate the assumed model.
- It should be reliable and reasonably efficient under small deviations from the
  model (e.g. when the distribution lies in a neighborhood of the assumed model).
- Larger deviations from the model should not affect the estimation procedure
  excessively.

A robust estimation method is a compromise with respect to these three features.
This compromise is illustrated by @anscombe1960rejection using an insurance
metaphor: "sacrifice some efficiency at the model in order to insure against
accidents caused by deviations from the model".

It is often believed that robust procedures may be avoided by using the 
following two-step procedure:

- Clean the data using some rule for outlier rejection.
- Apply classical optimal procedures on the "clean" data. 

Unfortunately such procedures cannot replace robust methods as discussed in
 @huber2009robust for the following reasons:

- It is rarely possible to seperate the two steps. For example, in a parametric
  regression setting, outliers are difficult to recognize without reliable 
  (i.e. robust) estimates of the model's parameters.
-  The cleaned data will not correspond to the assumed model since there will be
  statistical errors of both kinds (false acceptance and false rejection). 
  Therefore in general the classical theory is not applicable to the cleaned 
  sample.
- Empirically, the best rejection procedures do not reach the performance of 
  the best robust procedures. The latter are apparently superior because they 
  can make a smoother transition between the full acceptance and full rejection
  of an observation using weighting procedures @hampel1987robust.
- Empirical studies have also shown that many of the classical rejection methods
  are unable to deal with multiple outliers. Indeed, it is possible that a
  second outlier "masks" the effect of the first so that neither are rejected.

Unfortunately the least squares estimator suffers from a dramatic lack of
robustness. A single outlier can have an arbitrarily large effect on the
estimated parameters. In order to assess the robustness of an estimator we first
need to introduce an important concept, namely the influence function. This 
concept was introduced in @hampel1968contributions and it formalizes the bias
caused by one outlier. The influence function of an estimator represents the
effect of an infinitesimal contamination at the point $x$ or ($\mathbf{x}$, $y$)
in the regression setting) on the estimate, standardized by the mass of
contamination. Mathematically, the influence function of the estimator $T$ for
the model $F$ is given by:

\[\text{IF}(x| T, F) = \lim_{\varepsilon \rightarrow 0} \frac{T\left((1-\varepsilon) F + \varepsilon \Delta_x \right) - T\left(F\right)}{\varepsilon}\]

where $\Delta_x$ is a probability measure which puts mass $1$ at the point $x$.	

## The Classical Least-Squares Estimator

The standard definition of the linear model is derived as follows. 
Let ${(\mathbf{x}_{i},y_{i}): i = 1, \ldots, n}$ be a sequence of independent 
identically distributed random variables such that:

\[y_{i} = \mathbf{x}_{i}^{T} {\boldsymbol{\beta}} + u_{i}\]

where $y_{i} \in \mathbb{R}$ is the $i$-th observation, $\mathbf{x}_{i} \in \mathbb{R}^{p}$ 
is the $i$-th row of the design matrix $\mathbf{X} \in \mathbb{R}^{n\times p}$, 
$\boldsymbol{\beta} \in \boldsymbol{\Theta} \subseteq \mathbb{R}$ is a *p*-vector of unknown
parameters, $u_{i} \in \mathbb{R}$ is the $i$-th error.   

The least-squares estimator $\hat{\boldsymbol{\beta}}_{LS}$ of $\boldsymbol{\beta}$ can be
expressed as an $M$-estimator^[$M$-estimators are obtained as the
minima of sums of functions of the data.] Least-squares estimators are an
example of the larger class of $M$-estimators. The definition of $M$-estimators
was motivated by robust statistics which delivered new types of $M$-estimators.
defined by the estimating equation:

\begin{equation}
  \sum_{i = 1}^{n} \left(y_{i} - \mathbf{x}_{i}^{T} \boldsymbol{\beta} \right)\mathbf{x}_{i} = 0.
	(\#eq:lsMestim2)
\end{equation}

This estimator is optimal under the following assumptions:

- $u_{i}$ are normally distributed.
- $\mathbb{E}[u_{i}] = 0$ for $i = 1, \ldots, n$.
- $Cov(u_{1}, \ldots, u_{n}) = \sigma^2 \, \mathbf{I}_{n}$ where $\mathbf{I}_{n}$
  denotes the identity matrix of size $n$.

In other words, least-squares estimation is only optimal when the errors are
normally distributed. Small departures from the normality assumption for the
errors results in considerable loss of efficiency of the least-squares estimator
(see @hampel1987robust, @huber1973robust and @hampel1973robust).

## Robust Estimators for Linear Regression Models

The "Huber estimator" introduced in @huber1973robust was one of the first robust
estimation methods applied to linear models. Basically, this estimator is a
weighted version of the least-squares estimate with weights of the form:

\[
w_{i} = \min \left(1,\frac{c}{|r_{i}|}\right)
\]

where $r_{i}$ is the $i$-th residual and $c$ is a positive constant which
controls the trade-off between robustness and efficiency.

Huber proposed an \textit{M}-estimator $\hat{\boldsymbol{\beta}}_{H}$ of $\boldsymbol{\beta}$ 
defined by the estimating equation:

\[
	\sum_{i = 1}^{n} \psi_{c}\left(y_{i} - \mathbf{x}_{i}^{T} \boldsymbol{\beta} \right)\mathbf{x}_{i} = 0
\]

where $\psi_{c}(\cdot)$ corresponds to Huber's weight function 

\begin{equation}
w\left({x}\right) = \begin{cases}
1, &\text{if } \left|{x}\right| \le k \\
\frac{k}{\left|{x}\right|}, &\text{if} \left|{x}\right| > k
\end{cases}
(\#eq:huberweight)
\end{equation}

and, thus, is defined as:

\[
\psi_{c}(r) = \left\{  
\begin{array}{l l}
  r & \quad \\
  c \cdot \text{sign}(r). & \quad \\
\end{array} \right. 
\]

However, the Huber estimator cannot cope with problems caused by outlying points
in the design (or covariate) matrix $X$. An estimator which was developed to
address this particular issue is the one proposed by Mallows which has the 
important property that the influence function is bounded also for the matrix
$X$ (see @krasker1980estimation for more details).

## Applications of Robust Estimation

Having rapidly highlighted the theory of robustness, we now focus on the 
application of robust techniques in practical settings. Over the next three 
examples, estimation between classical or traditional methods will be compared
with  robust methods to illustrate the usefulness of using the latter techniques
can have in specific scenarios. 

```{example, label = "slmrobust"}

Consider a simple linear model with Gaussian errors such as

\begin{equation}
		y_i = \alpha + \beta x_i + \varepsilon_i, \;\; \varepsilon_i \overset{iid}{\sim} \mathcal{N}(0,\sigma^2)
  (\#eq:exam)
\end{equation}

for $i = 1,...,n$. Next, we set the parameter values $\alpha = 5$, 
$\beta = 0.5$ and $\sigma^2 = 1$ in order to simulate 20 observations from
the above simple linear model where we define $x_i = i$ for $i = 1,...,20$. In the left panel of Figure \@ref(fig:slmrobex), we present the simulated observations 
together with the fitted regression lines obtained by least-squares and a
robust estimation method. It can be observed that both lines are very similar
and "close" to the true model given by $y_i = 5 + 0.5 i$. Indeed, although the
robust estimator pays a small price in terms of efficiency compared to the the
least-squares estimator, the two methods generally deliver very "similar" results
when the model assumption holds. On the other hand, the robust estimators provide 
(in general) far more reliable results when outliers are present in a data-set.
To illustrate this behavior we modify the last observation by 
setting $\varepsilon_{20} = -10$ (which is "extreme" under the assumption 
that $\varepsilon_i \overset{iid}{\sim} \mathcal{N}(0,1)$). 
The modified observations are presented in the right panel of 
Figure \@ref(fig:slmrobex) together with the fitted regression lines. 
In this case, the least-squares is strongly influenced by the outlier 
we introduced while the robust estimator remains stable and "close" to the
true model.
```


```{r slmrobex, cache = TRUE, fig.cap="Simulation Study Comparing Robust and Classical Regression Methodologies", fig.width = 13}
# Load robust library
library("robustbase")

# Set seed for reproducibility
set.seed(867)

# Sample size
n = 20      

# Model's parameters
alpha = 5    # Intercept 
beta = 0.5   # Slope 
sig2 = 1     # Residual variance

# Construct response variable y
x = 1:n 
y = alpha + beta*x + rnorm(n,0,sqrt(sig2))

# Construct "perturbed" verison of y
y.perturbed = y
y.perturbed[20] = alpha + beta*x[20] - 10

# Compute LS estimates
LS.y = lm(y ~ x)
LS.y.fit = coef(LS.y)[1] + coef(LS.y)[2]*x
LS.y.pert = lm(y.perturbed ~ x)
LS.y.pert.fit = coef(LS.y.pert)[1] + coef(LS.y.pert)[2]*x

# Compute robust estimates
RR.y = lmrob(y ~ x)
RR.y.fit = coef(RR.y)[1] + coef(RR.y)[2]*x
RR.y.pert = lmrob(y.perturbed ~ x)
RR.y.pert.fit = coef(RR.y.pert)[1] + coef(RR.y.pert)[2]*x

# Define colors
gg_color_hue <- function(n, alpha = 1) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100, alpha = alpha)[1:n]
}
couleurs = gg_color_hue(6)

# Compare results based on y and y.perturbed
par(mfrow = c(1,2))
plot(NA, ylim = range(cbind(y,y.perturbed)), xlim = range(x),
     main = "Uncontaminated setting", xlab = "x", ylab = "y")
grid()
points(x,y, pch = 16, col = couleurs[4])
lines(x, alpha + beta*x, col = couleurs[3], lty = 2)
lines(x, LS.y.fit, col = couleurs[5])
lines(x, RR.y.fit, col = couleurs[1])

legend("topleft",c("Observations","Estimated model (least-squares)",
                   "Estimated model (robust)", "True model"), 
       lwd = c(NA,1,1,1), col = couleurs[c(4,5,1,3)],
       lty = c(NA,1,1,2), bty = "n", pch = c(16,NA,NA,NA))

plot(NA, ylim = range(cbind(y,y.perturbed)), xlim = range(x),
     main = "Contaminated setting", xlab = "x", ylab = "y")
grid()
points(x[1:19],y.perturbed[1:19], pch = 16, col = couleurs[4])
points(x[20], y.perturbed[20], pch = 15, col = couleurs[6])
lines(x, alpha + beta*x, col = couleurs[3], lty = 2)
lines(x, LS.y.pert.fit, col = couleurs[5])
lines(x, RR.y.pert.fit, col = couleurs[1])

legend("topleft",c("Uncontamined observations", "Contamined observation",
                   "Estimated model (least-squares)", "Estimated model (robust)",
                   "True model"), lwd = c(NA,NA,1,1,1), col = couleurs[c(4,6,5,1,3)],
       lty = c(NA,NA,1,1,2), bty = "n", pch = c(16,15,NA,NA,NA))
```

```{example, label="bslmrob", name = "Robust v. Classical Simulation Study."}
The next example presents a simulation study where the robust and classical
techniques will be compared in order to show that the previous example was not 
due to a "lucky" sample favorable to the robust approach. To do so, the 
simulation study will generate 100 realizations from the model in the previous 
example and use the same robust and classical estimators to retrieve the 
parameters for each of the 100 iterations. 
```

```{r lmbsrob, cache = TRUE, warning = FALSE, message = FALSE}
# Number of bootstrap replications
B = 10^3

# Initialisation
coef.LS.cont = coef.rob.cont = matrix(NA,B,2)
coef.LS.uncont = coef.rob.uncont = matrix(NA,B,2)

# Start Monte-carlo
for (j in 1:2){
  for (i in seq_len(B)) {
   # Control seed
    set.seed(2*j*B + i)
    
    # Uncontaminated case
    if (j == 1){
      y = alpha + beta*x + rnorm(n,0,sqrt(sig2))
      coef.LS.uncont[i,] = as.numeric(lm(y ~ x)$coef)
      coef.rob.uncont[i,] = as.numeric(lmrob(y ~ x)$coef)
    }
    
    # Contaminated case
    if (j == 2){
      y = alpha + beta*x + rnorm(n,0,sqrt(sig2))
      y[20] = alpha + beta*x[20] - 10
      coef.LS.cont[i,] = as.numeric(lm(y ~ x)$coef)
      coef.rob.cont[i,] = as.numeric(lmrob(y ~ x)$coef)
    }
  }
}

# Make graph
colors = gg_color_hue(6, alpha = 0.5)
names = c("LS - uncont","Rob - uncont","LS - cont","Rob - cont")

par(mfrow = c(1,2))
boxplot(coef.LS.uncont[,1],coef.rob.uncont[,1],coef.LS.cont[,1],
        coef.rob.cont[,1], main = expression(alpha), 
        col = colors[c(5,1,5,1)], 
        cex.main = 1.5, xaxt = "n")
axis(1, labels = FALSE)
text(x = seq_along(names), y = par("usr")[3] - 0.15, srt = 45, adj = 1,
     labels = names, xpd = TRUE)

abline(h = alpha, lwd = 2)

boxplot(coef.LS.uncont[,2],coef.rob.uncont[,2],coef.LS.cont[,2],
        coef.rob.cont[,2], main = expression(beta), 
        col = colors[c(5,1,5,1)], 
        cex.main = 1.5, xaxt = "n")
axis(1, labels = FALSE)
text(x = seq_along(names), y = par("usr")[3] - 0.015, srt = 45, adj = 1,
     labels = names, xpd = TRUE)
abline(h = beta, lwd = 2)
```

It can be seen that, as underlined earlier, that the estimations resulting from
the two methods appear to be quite similar, with the robust estimator performing
slightly less efficiently than the classical estimator. However, under the
contaminated setting it is evident how the robust estimation procedure is not 
affected much by the outliers in the simulated data while the classical 
techniques appear to be highly biased. Therefore, if there appear to be outliers
in the data, a robust estimation procedure may considered preferable. In fact, 
the results of the two estimations could be compared to understand if there
appears to be some deviation from the model assumptions.

```{example, dsbook}
A practical example which underlines the usefulness of robust estimation
procedures is given by the dataset that contains the properties of 47 stars
from the Hertzsprung-Russell diagram of the star cluster CYG OB1 in the
direction of Cygnus. From the plot it can be observed that there appears to be
a cluster of four stars on the upper left hand-side of the plot. The rest of
the data however appears to have a reasonably good linear behavior.
```

```{r robstarcluster, cache = TRUE, fig.cap="Comparison of Estimation Methodologies on 47 observations from the Hertzsprung-Russell diagram of the star cluster CYG OB1 in the direction of Cygnus."}
colors = gg_color_hue(6)
data(starsCYG, package = "robustbase")
par(mfrow = c(1,1))
plot(NA, xlim = range(starsCYG$log.Te) + c(-0.1,0.1),
     ylim = range(starsCYG$log.light),
     xlab = "Temperature at the surface of the star (log)",
     ylab = "Light intensity (log)")
grid()  
points(starsCYG, col = colors[4], pch = 16)

LS = lm(starsCYG$log.light ~ starsCYG$log.Te)
rob = lmrob(starsCYG$log.light ~ starsCYG$log.Te)
x = seq(from = min(starsCYG$log.Te)-0.3, 
        to = max(starsCYG$log.Te)+0.3, length.out = 4)
fit.LS = coef(LS)[1] + x*coef(LS)[2]
fit.rob = coef(rob)[1] + x*coef(rob)[2]

lines(x, fit.LS, col = colors[5])
lines(x, fit.rob, col = colors[1])


legend("topright",c("Observations","Estimated model (least-squares)",
                    "Estimated model (robust)"), lwd = c(NA,1,1), 
       col = colors[c(4,5,1)], lty = c(NA,1,1), 
       bty = "n", pch = c(16,NA,NA))

```

From Figure \@ref(fig:robstarcluster) it can be seen that classical linear
regression is considerably affected by the cloud of apparent outliers on the
left hand side of the plot. As a result, the regression line has a negative 
slope when the majority of the data appears to have a clear positive linear
trend. The latter is however the case when using a robust approach which 
indeed detects a positive linear trend. Having said this, there are in fact
two distinct populations in the data consisting in "giants" (upper left corner) 
and "main sequencers" (right handside). Thus, when observing data in general it 
is always necessary to understand if we are dealing with real 
outliers (unexplained outlying data) or simply with data that can be explained
by other factors (e.g. a different population as in this example).

```{r estimatesreg, cache = TRUE}
summary(LS)
summary(rob)
```
